% ---------------------------------------------------------------------
% ---------------------------------------------------------------------
% ---------------------------------------------------------------------

\chapter*{\abstractname}
\markboth{Abstract}{}
% ---------------------------------------------------------------------
% ---------------------------------------------------------------------
% ---------------------------------------------------------------------


Reduced-order models (ROMs) have become indispensable tools for reducing the computational complexity of high-fidelity simulations in science and engineering.  In this thesis, we introduce a novel training framework that combines the integral form of Operator Inference (OpInf) with adjoint-state methods to yield robust, data-driven ROMs. By formulating a continuous-time loss functional that integrates the governing equations over time, our approach avoids explicit differentiation of noisy state data and inherently regularizes the learning problem. We derive the corresponding adjoint equations for exact gradient computation of the loss functional and present a gradient-descent minimization algorithm that updates ROM parameters efficiently.

We validate the proposed method on two canonical nonlinear PDEs: the viscous Burgers’ equation and the two-dimensional Fisher-KPP reaction-diffusion equation. In systematic numerical experiments, we compare the adjoint-trained ROM against the standard OpInf approach under (i) varying snapshot density and (ii) additive Gaussian noise. Our results show that both methods perform comparably when snapshots are uniformly sub-sampled, but the adjoint method exhibits improved accuracy and stability in the presence of noisy data.  Moreover, although each training iteration requires a backward pass, the overall computational cost scales constantly with the number of parameters, making the method favorable for very high-dimensional full models.



  




